---
title: "Recommendations for existing bikeshare subscription models"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE)
```
##### Aims: Study current subscription model feasibility of Blue Bikes(Boston) and Divvybikes (Chicago) to drive acquisition of new subscribers.

```{r libraries, include=FALSE}
library(ggplot2)
library(tidyr)
```
##### 1. (2017-2019) Trip Counts from Divvybikes and Bluebikes

```{r q1-biketrips, include=FALSE}
q1 <- read.csv("https://raw.githubusercontent.com/cheryltky/bikeshare_studies/main/csvdata/Q1.csv", sep=",", header = T)
head(q1)
str(q1)
#add index to dataframe, remove NAs
q1$index <- seq_len(dim(q1)[1])
q1 <-na.omit(q1)
q1
```


```{r divvy-blue, echo = FALSE}

plot(q1$index, q1$trips_bluebike, pch=20, col="blue", bty='n', xaxt='n', xlab='', ylab="Number of Trips", ylim=c(min(c(q1$trips_bluebike, q1$trips_divvy)), max(c(q1$trips_bluebike, q1$trips_divvy))))
points(q1$trips_divvy, pch=20, col='red')
axis(1, at=q1$index, labels=q1$month_year, las=2)
lines(q1$index, q1$trips_bluebike, col="blue", lwd=1)
lines(q1$index, q1$trips_divvy, col="red", lwd=1)

```


```{r ggplot method, include=FALSE}
dat <- read.csv("https://raw.githubusercontent.com/cheryltky/bikeshare_studies/main/csvdata/Q1.csv", sep=",", header = T)

dat$index <- seq_len(dim(dat)[1])
datl <- gather(dat, company, number_of_trips, trips_bluebike:trips_divvy)

datl

ggplot(datl, aes(x=index, y=number_of_trips, color=as.factor(company))) + geom_point() + scale_x_continuous(breaks=dat$index, labels=dat$month_year) + theme(axis.text.x = element_text(angle=90)) + xlab("Date") + ylab("Number of Trips") + geom_line(data = datl, aes(x=index, y=number_of_trips))

geom_smooth(method="loess")
str(datl)
```

##### 2. Subscription Segment Breakdown
```{r subscriptions, echo=FALSE, results=FALSE}
dat2 <- read.csv("https://raw.githubusercontent.com/cheryltky/bikeshare_studies/main/csvdata/q2.csv", fileEncoding="UTF-8-BOM", na.strings = '..',sep = ",", header=T)

dat2 <- na.omit(dat2)
dat2
head(dat2)
str(dat2)

library(scales)
#barplot
ggplot(dat2,aes(x= year,y=user_count, fill=segment)) + geom_bar(position ="dodge", stat="identity")+
  scale_y_continuous(
  labels = scales::comma_format(big.mark = ','))

```

##### Q3. Seasonal vs Daily Commuter Frequencies

```{r-seasonaltrends}

```

##### Q4. What was the longest journey and what do we know about it?
##### Q5. How often do bikes need to be relocated?
##### Q6. How far is a typical journey?
